What Makes Us Happy? Examining the Impact of Social, Economic, and Personal Factors

### 2024-12-03 paper/paper.qmd
make some changes

# Data {#sec-data}

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-happiness_preview
#| tbl-cap: "Happiness Data Sample Preview"

@@

## variables

@@

```{r}
#| echo: false
#| warning: false
#| label: fig-Marital
#| fig-cap: Happiness Levels Across Different Marital Statuses

@@

@fig-Marital illustrates the relationship between marital status and happiness levels, with data points grouped by marital categories for clarity. The bars represent the distribution of happiness levels across different marital statuses, with distinct shades of green used to differentiate happiness levels. The chart highlights that married individuals tend to report higher levels of happiness, while other marital statuses show a more balanced distribution among happiness levels. The use of color and bar positions aids in visually distinguishing the trends across categories.

@@

```{r}
#| echo: false
#| warning: false
#| label: fig-Childs
#| fig-cap: Distribution of the Number of Children by Happiness Level

@@

# Create a boxplot
ggplot(data, aes(x = happy, y = childs, fill = happy)) +
  geom_boxplot(outlier.color = "red", outlier.size = 1.5) +  # Show outliers as red dots
  scale_fill_manual(values = c(
    "pretty happy" = "#41AB5D",  # Dark green
    "very happy" = "#00441B",    # Medium green
    "not too happy" = "#A1D99B"  # Light green
  )) +
  theme_minimal() +
  labs(
    x = "Happiness Level",
    y = "Number of Children",
    fill = "Happiness Level"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
    plot.title = element_text(face = "bold", hjust = 0.5)  # Bold and center the title
  )

```
@fig-Childs illustrates the relationship between the number of children and happiness levels, with data points grouped by happiness categories for clarity. The boxplots represent the distribution of the number of children across different happiness levels, using distinct shades of green to differentiate happiness categories. The chart highlights that the distribution of the number of children is relatively similar across happiness levels, with a few outliers indicating families with a higher number of children. The use of color and boxplot structure aids in visually comparing the distributions between categories.

@@

```{r}
#| echo: false
#| warning: false
#| label: fig-Degree
#| fig-cap: Proportion of Happiness Levels by Education Level

@@

```
@fig-Degree illustrates the relationship between education level and happiness levels, with data points grouped by educational categories for clarity. The bars display the proportion of happiness levels by education categories, using distinct shades of green to represent happiness levels. The chart indicates that individuals with higher education, such as graduate degrees, are more likely to report higher happiness levels, while other education categories exhibit more uniform distributions. Proportional stacking and color differentiation help compare trends effectively across education levels.

@@

```{r}
#| echo: false
#| warning: false
#| label: fig-Satjob
#| fig-cap: Proportion of Happiness Levels by Job Satisfaction

@@

```
@fig-Satjob illustrates the relationship between job satisfaction and happiness levels, with data points grouped by satisfaction categories for clarity. The bars display the proportion of happiness levels by job satisfaction categories, using distinct shades of green to represent happiness levels. The chart indicates that individuals with higher job satisfaction are more likely to report greater happiness, while those with lower satisfaction exhibit a broader distribution. Proportional stacking and color differentiation facilitate comparisons across job satisfaction levels.

@@

```{r}
#| echo: false
#| warning: false
#| label: fig-Realrinc
#| fig-cap: Real Income vs. Happiness Level

@@

```
@fig-Realrinc illustrates the relationship between real income and happiness levels, with data points grouped by happiness categories for clarity. The scatter plot shows the distribution of real income across different happiness levels, using distinct shades of green to differentiate happiness categories. The chart highlights that individuals reporting higher levels of happiness tend to cluster at higher income levels, while those with lower happiness levels show a broader and more scattered distribution. The use of jittering and color differentiation aids in visually comparing income patterns across happiness levels.

@@

@@

>>>>

# Data {#sec-data}
## Data Source

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-happiness_preview
#| tbl-cap: "Sample Preview of Happiness Data: A snapshot showing key variables"

@@

## Variables

@@

```{r}
#| echo: false
#| warning: false
#| label: fig-MaritalStatusDistribution
#| fig-cap: "Happiness Levels Across Different Marital Statuses: Depicting the Distribution of Self-Reported Happiness Levels ('Very Happy', 'Pretty Happy', 'Not Too Happy') by Marital Status Categories"

@@

@fig-MaritalStatusDistribution illustrates the relationship between marital status and happiness levels, with data points grouped by marital categories for clarity. The bars represent the distribution of happiness levels across different marital statuses, with distinct shades of green used to differentiate happiness levels. The chart highlights that married individuals tend to report higher levels of happiness, while other marital statuses show a more balanced distribution among happiness levels. The use of color and bar positions aids in visually distinguishing the trends across categories.

@@

```{r}
#| echo: false
#| warning: false
#| label: fig-ChildrenDistribution
#| fig-cap: "Distribution of the Number of Children by Happiness Level: Showing Family Size Across Different Happiness Categories"

@@

# Create a violin plot with jittered points
ggplot(data, aes(x = happy, y = childs, fill = happy)) +
  geom_violin(trim = TRUE, alpha = 0.7) +  # Violin plot to show distribution
  geom_jitter(width = 0.2, alpha = 0.4, color = "black", size = 1) +  # Add jitter points for raw data
  scale_fill_manual(values = c(
    "pretty happy" = "#41AB5D",  # Dark green
    "very happy" = "#00441B",    # Medium green
    "not too happy" = "#A1D99B"  # Light green
  )) +
  theme_minimal() +
  labs(
    x = "Happiness Level",
    y = "Number of Children",
    fill = "Happiness Level"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
    plot.title = element_text(face = "bold", hjust = 0.5),  # Bold and center the title
    legend.title = element_text(face = "bold"),  # Bold legend title
    panel.grid.major = element_blank(),  # Remove major gridlines
    panel.grid.minor = element_blank()   # Remove minor gridlines
  )

```
@fig-ChildrenDistribution illustrates the relationship between the number of children and happiness levels, with data points grouped by happiness categories for better clarity. The violin plots display the density distribution of the number of children across the three happiness levels: "Not Too Happy," "Pretty Happy," and "Very Happy," differentiated by distinct shades of green. Additionally, jittered points overlay the violin plots, representing individual observations, thereby providing a detailed view of the raw data.

The chart highlights that the number of children is distributed similarly across happiness levels, with the highest density of observations clustering around smaller family sizes. A few data points stand out as outliers, representing families with larger numbers of children. This combination of violin plots and jittered points effectively showcases both the overall trends and individual variances, making the relationships between family size and happiness levels visually accessible and easy to interpret.

@@

```{r}
#| echo: false
#| warning: false
#| label: fig-EducationHappinessDistribution
#| fig-cap: "Proportion of Happiness Levels by Education Level: Representing the Distribution of Happiness Categories Across Different Levels of Educational Attainment"

@@

```
@fig-EducationHappinessDistribution illustrates the relationship between education level and happiness levels, with data points grouped by educational categories for clarity. The bars display the proportion of happiness levels by education categories, using distinct shades of green to represent happiness levels. The chart indicates that individuals with higher education, such as graduate degrees, are more likely to report higher happiness levels, while other education categories exhibit more uniform distributions. Proportional stacking and color differentiation help compare trends effectively across education levels.

@@

```{r}
#| echo: false
#| warning: false
#| label: fig-JobSatisfactionHappiness
#| fig-cap: "Proportion of Happiness Levels by Job Satisfaction: Depicting the Distribution of Self-Reported Happiness Categories Across Different Job Satisfaction Levels"

@@

```
@fig-JobSatisfactionHappiness illustrates the relationship between job satisfaction and happiness levels, with data points grouped by satisfaction categories for clarity. The bars display the proportion of happiness levels by job satisfaction categories, using distinct shades of green to represent happiness levels. The chart indicates that individuals with higher job satisfaction are more likely to report greater happiness, while those with lower satisfaction exhibit a broader distribution. Proportional stacking and color differentiation facilitate comparisons across job satisfaction levels.

@@

```{r}
#| echo: false
#| warning: false
#| label: fig-IncomeHappiness
#| fig-cap: "Real Income and Happiness Level: This analysis explores the distribution of real income across different happiness categories, using jitter points to illustrate the variability of individual data within each category"

@@

```
@fig-IncomeHappiness illustrates the relationship between real income and happiness levels, with data points grouped by happiness categories for clarity. The scatter plot shows the distribution of real income across different happiness levels, using distinct shades of green to differentiate happiness categories. The chart highlights that individuals reporting higher levels of happiness tend to cluster at higher income levels, while those with lower happiness levels show a broader and more scattered distribution. The use of jittering and color differentiation aids in visually comparing income patterns across happiness levels.

@@

## Model Limitations

Although the Bayesian logistic regression model performs well in analyzing the relationship between happiness and its predictors, it still has certain limitations. First, the model assumes a linear relationship between the predictors and the log-odds of happiness, which may overlook complex non-linear or interaction effects between variables. For example, the relationship between age and happiness may exhibit non-linear patterns, leading to potential biases in estimation. Second, the imbalanced distribution of categories in the data—where the "very happy" category has fewer samples—may cause the model to favor predicting the majority classes (e.g., "not too happy" and "pretty happy") while underperforming in classifying the minority class. Lastly, the model employs weakly informative priors (e.g., Normal(0, 2) and Cauchy(0, 2)) to ensure flexibility. However, in cases with small samples or rare categories, these priors may impact the stability of coefficient estimates, increasing uncertainty in the results.

## Alternative Models Considered

The Bayesian logistic regression model was chosen due to its unique ability to incorporate prior knowledge, quantify uncertainty, and provide robust parameter estimates in the analysis of complex individual-level happiness data. Unlike traditional logistic regression, the Bayesian framework allows for the inclusion of weakly informative priors, which help stabilize estimates in the presence of limited or imbalanced data (e.g., the "very happy" category being underrepresented) while avoiding overfitting. This is particularly advantageous in handling imbalanced target variable distributions, where the Bayesian model performs better in classifying minority categories. Additionally, the Bayesian model computes posterior distributions for each predictor, offering a comprehensive understanding of the uncertainty and variability in the estimated effects. This probabilistic framework not only enhances interpretability but also supports better decision-making by providing credible intervals instead of single-point estimates.

In contrast, other models such as decision trees may perform poorly when analyzing complex social data. While decision trees offer advantages in interpretability and handling non-linear relationships, they are highly sensitive to changes in data distribution and noise. In cases of class imbalance, decision trees tend to overfit the majority class, reducing their predictive accuracy for underrepresented categories. Moreover, decision trees lack the ability to provide posterior distributions or quantify uncertainty, making it challenging to robustly evaluate the effects of predictors. These limitations render decision trees less reliable for analyzing complex problems like happiness data.

For these reasons, the Bayesian logistic regression model stands out for its flexibility and robustness, capturing complex non-linear relationships and potential interactions while quantifying uncertainty. It offers more accurate and comprehensive insights for predicting happiness compared to alternative models, making it a superior choice for this analysis.

@@

# Data cleaning process

To prepare the dataset for analysis, I conducted a thorough cleaning process to address missing values, inconsistent formatting, and invalid entries, ensuring the data’s accuracy and usability. The cleaning steps were designed to align with the study’s focus on socio-economic and personal factors influencing happiness.

## Handling Missing and Invalid Values

- In the `realrinc` (real income) column, entries coded as `-100` were replaced with `NA` to signify missing data, as `-100` was an invalid placeholder value.

- For categorical variables like `marital`, `degree`, `satjob`, and `happy`, entries such as `.n: no answer` and `.d: do not know/cannot choose` were converted to `NA` to exclude non-informative responses. Additionally, in the `satjob` column, responses marked as `.i: inapplicable` were also replaced with `NA` to ensure only relevant data remained.

## Standardizing Categorical Data

- Categorical variables, including `marital`, `degree`, `sex`, `happy`, `satjob`, and `ballot`, were cleaned by removing extra spaces and converting all text to lowercase to maintain consistency and facilitate accurate analysis.

## Adjusting the `childs` Column

- Entries labeled as `"8 or more"` in the `childs` (number of children) column were replaced with `8`, as this was the maximum numeric representation available. This transformation allowed the variable to be converted into a numeric format for quantitative analysis.

## Converting Data Types
- Variables like `age` and `childs` were converted to numeric data types to enable proper statistical analysis and modeling.

## Filtering Irrelevant Responses
- Rows where the `happy` variable contained responses like `.d: do not know/cannot choose` or `.n: no answer` were removed to ensure the dataset only included valid observations relevant to the study’s goal of understanding happiness.

## Summary
These cleaning steps were essential for maintaining data integrity and ensuring that the analysis would provide meaningful and reliable results. By standardizing and filtering the data, the study could focus on valid and interpretable responses, minimizing noise and maximizing the quality of insights derived from the dataset.
